---
title: Protected Natural Area location
author: Sergio Lopez Olvera
date: '2022-07-02'
slug: []
categories:
  - R
tags:
  - natural
  - area
  - ecology
  - gis
  - location
  - map
  - plot
  - ggplot2
  - cleaning
  - wrangling
  - parsing
  - mining
  - data
  - pdf
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(readxl)
library(readr)
library(pdftools)
library(stringr)
library(raster)
library(rgdal)
library(ggmap)
library(broom)
library(RColorBrewer)
library(rgeos)
library(sp)
library(sf)
```

# Data import

Imported a .pdf document from a mexican government website, splitted strings on "/n" and unlisted the items. Then detected the string "Página 50" to get to the page where the coordinates table begins. The document looks like below.

```{r import, echo=FALSE, message=FALSE, warning=FALSE,}
#Import file
txt <- pdf_text("http://repositorio.veracruz.gob.mx/medioambiente/wp-content/uploads/sites/9/2018/02/Archipi%C3%A9lago-BS-region-capital_2015.pdf")

txt1 <- str_split(txt, "\n") %>% unlist()
head(txt1)
```

# Data cleaning and wrangling
Cleaned and wrangled the data using string trimming, detecting and replacing into a tidy format tibble. The tibble below shows one of the polygons' coordinates

```{r wrangling, echo=FALSE, warning=FALSE, message=FALSE}
  quinto <- txt1[1569:1961]

  not_ind_quinto <- quinto %>%
    str_which("Página")
  quinto <- quinto[-c(not_ind_quinto)]
  
  ind_quinto <- str_which(quinto, "\\d+")
  quinto <- quinto[ind_quinto]
  
  quinto <- quinto %>%
    str_replace("    ", " ") %>%
    str_replace("   ", " ") %>%
    str_replace("  ", " ") %>%
    str_trim(side = "both")
  
  quinto <- quinto %>%
    str_split("\\s(?=\\d{1,3}\\s)") %>%
    unlist() %>%
    str_replace("    ", " ") %>%
    str_replace("   ", " ") %>%
    str_replace("  ", " ") %>%
    str_trim(side = "both")
  
  quinto <- quinto %>%
    str_split(" ", simplify = TRUE) %>%
    as.tibble
  
  quinto <- quinto %>%
    mutate(V1 = as.numeric(V1), V2 = as.double(V2), V3 = as.double(V3)) %>%
    arrange(V1) 
  
  quinto <- quinto[-(seq(2, 64, 2)),]
  quinto <- quinto[-(seq(465, 1000, 1)),]
  
  quinto_tidy <- quinto[,2:3] %>% filter(!is.na(V2)&!is.na(V3))
  
  write.csv(quinto_tidy, "C:/Users/Sergio/projects/pixquiac/data/ANP/cuarto_t.csv")
  
  head(quinto_tidy)
```

# Spatial polygons building

Then, coordinates in a tidy format where converted to spatial polygons. Such as the one shown below

```{r sp, echo=FALSE, warning=FALSE, message=FALSE}

 x_quinto <- quinto_tidy[2]
  y_quinto <- quinto_tidy[1]
  poly_quin <- sp::Polygon(cbind(x_quinto, y_quinto))
  poly_quinto <- sp::Polygons(list(poly_quin), ID = "A")
  sp_quin <- sp::SpatialPolygons(list(poly_quinto), proj4string=CRS("+proj=utm +zone=14 +datum=WGS84 +units=m +no_defs"))

sp_quin
```

# Visualizing spatial polygon in a map

Such polygons were projected onto a map to visualize their spatial distribution. The map below shows the location of one of the seven polygons of the protected area.

```{r map, echo=FALSE, message=FALSE, warning=FALSE}
  sp_quin %>%
    ggplot(aes(lat, long)) +
    geom_polygon(fill = "blue") +
    labs(title = "Natural Protected Area location",
         subtitle = "Fifth polygon",
         x = "latitude",
         y = "longitude")
```
